---
title: 'BCM Growth Trajectories: Mixed Effects Modeling Demo'
author: "D2K BCM Growth Trajectories Team"
date: "2023-11-07"
output: pdf_document
---

## This file contains the demo for constructing, running, checking,
## validating, and plotting the linear mixed-effects models predicting children's
## BMI based on child opportunity index (COI), artificial light at night (ALAN),
## and other child- and school-level characteristics.

## Necessary data files: student_melt_new (output from 'mixed_effects_cleaning_data.py'),
##  student_data_cleaned (output from 'clean_student_data_second.py'),
##  coi_data_cleaned (retrieve from data source)

## Necessary R version: >= R-4.1.3

## _____________________________________________________________________________

## ENVIRONMENT AND DATA SETUP

```{r}
# clear global environment
rm(list = ls())
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Import necessary packages and libraries.
```{r}
## run the following lines to install necessary packages
devtools::install_version("jtools", "2.2.0")
devtools::install_version("optimx", "2023-10.21")
devtools::install_version("lme4", "1.1-32")
devtools::install_version("merTools", "0.6.1")
devtools::install_version("coxed", "0.3.3")
devtools::install_version("kableExtra", "1.3.4")
devtools::install_version("performance", "0.10.8")
devtools::install_version("ggh4x", "0.2.6")
devtools::install_version("tidyverse", "1.3.2")
remotes::install_version("Rttf2pt1", version = "1.3.8")

# run the following lines to load packages
library(jtools)
library(optimx)
library(lme4)
library(merTools)
library(coxed)
library(kableExtra)
library(performance)
library(ggh4x)
library(tidyverse)
```

Import the data.
```{r}
student_data = read.csv("~/BCM_GrowthTrajectories_F23/data_directory/melt_final.csv")
## or choose file interactively with the following command
#student_melted = read.csv(file.choose())

student_full = read.csv("~/BCM_GrowthTrajectories_F23/data_directory/student_data_cleaned.csv")
## or choose file interactively with the following command
#student_full = read.csv(file.choose())

# COI data (for adding in original COI categories)
coi_data = read.csv("~/BCM_GrowthTrajectories_F23/data_directory/coi_data_cleaned.csv")
## or choose file interactively with the following command
#coi_data = read.csv(file.choose())
```

Import mixed-effects modeling functions from .R file
```{r}
source("mixed_effects_modeling.R")
```

________________________________________________________________________________

## CLEAN DATA

```{r}
# Include COI categories from coi_data into student_data
student_melted$COI_cat <- coi_data$c5_COI_nat[match(student_data$ctract, coi_data$census_tract)]
```

```{r}
# Apply data cleaning functions
student_melted_cleaned <- clean_training_data(student_melted, student_full)
lm_plot <- plot_samples(student_melted_cleaned)
#print(lm_plot)
```

________________________________________________________________________________

## FIT MODELS

```{r}
#fit models
coi_model <- fit_coi_model(student_melted_cleaned)
dmsp_model <- fit_dmsp_model(student_melted_cleaned)

print(model_summary(coi_model))
print(model_summary(dmsp_model))
```

________________________________________________________________________________

## ASSUMPTIONS CHECKING

```{r}
#check model assumptions
print(check_assumptions(coi_model))
```

________________________________________________________________________________

## BOOTSTRAP FOR MODEL ESTIMATES AND CONFIDENCE INTERVALS

Note: depending on number of simulations, each bootstrap may take multiple hours to run.
```{r}
#run bootstrap
coi_bootstrapping <- model_bootstrapping(coi_model, "COI", num_simulations = 500)
alan_bootstrapping <- model_bootstrapping(dmsp_model, "ALAN", num_simulations = 500)

#create dataframes of results
coi_coeff_df <- create_coeff_df(coi_bootstrapping, coi_model, "COI")
dmsp_coeff_df <- create_coeff_df(alan_bootstrapping, dmsp_model, "ALAN")

#plot results
coi_dw_plot <- create_dw_plot(coi_coeff_df, "COI")
alan_dw_plot <- create_dw_plot(dmsp_coeff_df, "ALAN")

#print resulting plots
print(coi_dw_plot)
print(alan_dw_plot)
```

________________________________________________________________________________

## PREDICTIONS FOR VISUALIATION

```{r}
#get predictor dataframe for hypothetical "mean/mode student"
coi_preds_df <- get_predictions_df(student_melted, "COI")
dmsp_preds_df <- get_predictions_df(student_melted, "ALAN")

#get point estimate predictions and prediction intervals for student
coi_point_preds <- get_point_predictions(coi_model, coi_preds_df)
dmsp_point_preds <- get_point_predictions(dmsp_model, dmsp_preds_df)

#display the predicted trajectories and prediction intervals for student
display_bmi_preds(coi_preds_df, coi_point_preds, "COI")
display_bmi_preds(dmsp_preds_df, dmsp_point_preds, "ALAN")
```

________________________________________________________________________________

# VALIDATION TESTING

Note: block cross-validation (in total) may take multiple hours to run.
```{r}
#run block cross-validation for model versions (including "full" model with all
#potential predictors, a model with no age^2 term, a model with no random slopes
#for age, and a model with neither an age^2 term nor random slopes for age.)
coi_full_valid <- apply_blockCV(input_data = student_data, 
                                input_model = bmi_model_coi_full)

coi_noage2_valid <- apply_blockCV(input_data = student_data, 
                                  input_model = bmi_model_coi_noage2)

coi_noageslopes_valid <- apply_blockCV(input_data = student_data, 
                                       input_model = bmi_model_coi_noageslopes)

coi_noage2_noageslopes_valid <- apply_blockCV(input_data = student_data, 
                                 input_model = bmi_model_coi_noage2_noageslopes)

alan_full_valid <- apply_blockCV(input_data = student_data, 
                                 input_model = bmi_model_alan_full)

alan_noage2_valid <- apply_blockCV(input_data = student_data, 
                                   input_model = bmi_model_alan_noage2)

alan_noageslopes_valid <- apply_blockCV(input_data = student_data, 
                                        input_model = bmi_model_alan_noageslopes)

alan_noage2_noageslopes_valid <- apply_blockCV(input_data = student_data, 
                                input_model = bmi_model_alan_noage2_noageslopes)

# collect validation results for ALAN models  
alan_valid <- data.frame(full_rmse = alan_full_valid[1,],
                         full_mae = alan_full_valid[2,],
                         noage2_rmse = alan_noage2_valid[1,],
                         noage2_mae = alan_noage2_valid[2,],
                         noageslopes_rmse = alan_noageslopes_valid[1,],
                         noageslopes_mae = alan_noageslopes_valid[2,],
                         noage2_noageslopes_rmse = alan_noage2_noageslopes_valid[1,],
                         noage2_noageslopes_mae = alan_noage2_noageslopes_valid[2,])

#collect validation results for COI models
coi_valid <- data.frame(full_rmse = coi_full_valid[1,],
                        full_mae = coi_full_valid[2,],
                        noage2_rmse = coi_noage2_valid[1,],
                        noage2_mae = coi_noage2_valid[2,],
                        noageslopes_rmse = coi_noageslopes_valid[1,],
                        noageslopes_mae = coi_noageslopes_valid[2,],
                        noage2_noageslopes_rmse = coi_noage2_noageslopes_valid[1,],
                        noage2_noageslopes_mae = coi_noage2_noageslopes_valid[2,])

#calculate averages across cross-validation iterations
round(coi_valid %>% apply(MARGIN = 2, FUN = mean),4)
round(alan_valid %>% apply(MARGIN = 2, FUN = mean),4)
```

Calculate AIC and BIC for COI and ALAN models.
```{r}
AIC_COI <-
  c(
    AIC(bmi_model_coi_full(student_data)),
    AIC(bmi_model_coi_noage2(student_data)),
    AIC(bmi_model_coi_noageslopes(student_data)),
    AIC(bmi_model_coi_noage2_noageslopes(student_data)))

BIC_COI <- 
  c(
    BIC(bmi_model_coi_full(student_data)),
    BIC(bmi_model_coi_noage2(student_data)),
    BIC(bmi_model_coi_noageslopes(student_data)),
    BIC(bmi_model_coi_noage2_noageslopes(student_data)))

AIC_ALAN <-
  c(
    AIC(bmi_model_alan_full(student_data)),
    AIC(bmi_model_alan_noage2(student_data)),
    AIC(bmi_model_alan_noageslopes(student_data)),
    AIC(bmi_model_alan_noage2_noageslopes(student_data)))

BIC_ALAN <-
  c(
    BIC(bmi_model_alan_full(student_data)),
    BIC(bmi_model_alan_noage2(student_data)),
    BIC(bmi_model_alan_noageslopes(student_data)),
    BIC(bmi_model_alan_noage2_noageslopes(student_data)))

#collect AIC and BIC results
AIC_BIC <- data.frame(AIC_ALAN, AIC_COI, BIC_ALAN, BIC_COI)
```


