---
title: 'BCM Growth Trajectories: Data Cleaning, Step 1'
author: "BCM D2K Growth Trajectory Team"
date: "2023-09-14"
output: pdf_document
---

This file is the first step in cleaning student data, merging together
student data, school-census tract data, DMSP data, and COI data, as well as 
running  the CDC package for percentiles and BMI scores, screening extreme and 
error values, renaming columns, initializing data structures for use in R, and 
checking and visualizing missing data proportions. For access to sensitive data, 
see the project's Box file.

Necessary R version: >= R-4.1.3

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

________________________________________________________________________________


# Environment and Data Setup

```{r, warning = FALSE}
#clear global environment
rm(list = ls())

## run the following lines to install necessary packages
devtools::install_version("readxl", "1.4.1")
devtools::install_version("tidyverse", "1.3.2")
devtools::install_version("data.table", "1.14.6")
devtools::install_version("cdcanthro", "0.1.1")
devtools::install_version("measurements", "1.5.1")
devtools::install_version("forcats", "0.5.2")
devtools::install_version("lubridate", "1.9.0")

#run the following lines to load packages
library(readxl) #version 1.4.1
library(tidyverse) #version 1.3.2
library(data.table) #version 1.14.6
library(cdcanthro) #version 0.1.1
library(measurements) #version 1.5.1
library(forcats) #version 0.5.2
library(lubridate) #version 1.9.0
```

```{r, warning = FALSE}
#set working directory
working_directory <- "../data_directory/"
setwd(work_dir)

#read in .txt file for file paths (can edit file to match own file paths)
file_paths <- read.delim("file_paths.txt")

#read in data
growth_init <- read.csv(file_paths$file_path[1])
tracts <- read_xlsx(file_paths$file_path[2])
DMSP <- read.csv(file_paths$file_path[3])
COI <- read.csv(file_paths$file_path[4])
```

________________________________________________________________________________


# Initial Data Cleaning

```{r}
#character strings to date format, using standard US date structure
toDate <- function(x){as.Date(x, format = "%m/%d/%Y")}

##do initial cleaning (select and homogenize columns)
growth <- growth_init %>% 
  select(`ï..id`, #student id
         ethnic, #ethnicity
           #1 white, 2 Black, 3 Hispanic, 4 Asian, 5 Indigenous American
         gender,
           #1 male, 2 female; matches CDC coding
         dob, #date of birth
         starts_with("dcoll_"), #date collected
         starts_with("ht_"), #height at collection pt X
           #in inches (based on context)
         starts_with("wt_"), #weight at collection pt X
           #in lbs (based on context)
         starts_with("BMI_"), #calculated BMI
         contains("chool_"), #school at collection pt X
         starts_with("cdcage_"), #age in months at collection pt X, CDC definition
         ) %>%
  #standardize naming schema
  rename(id = `ï..id`,
         school_9 = School_9,
         sex = gender) %>% #gender to sex for use with CDC functions
  #clean and change variable types
  mutate(dob = toDate(dob),
         across(dcoll_1:dcoll_11, toDate),
         id = as.character(id),
         ethnic = as.factor(ethnic),
         sex = as.factor(sex),
         across(school_1:school_11, as.factor))

#head(growth)
colnames(growth)
#summary(growth)
```

##Restructure data
```{r}
#create list of dataframes so we can use vectorized methods
#each list element is a variable: static vars have 1 column, non-static have 11
growth_list_init <- list(static = growth %>% select(id, ethnic, sex, dob),
                    dcoll = growth %>% select(starts_with("dcoll_")),
                    ht_in = growth %>% select(starts_with("ht_")),
                    wt_lbs = growth %>% select(starts_with("wt_")),
                    BMI = growth %>% select(starts_with("BMI_")),
                    school = growth %>% select(starts_with("school_")),
                    cdcage = growth %>% select(starts_with("cdcage_"))
                    )

names(growth_list_init)
```

##Clean data generated in error
```{r}
#check for negative changes in height
#which(growth_list_init$chght < 0)

#function: adjust "diff" function to ignore NAs
  #input: vector (or row or column of dataframe)
  #output: vector of same length with differences between subsequent values,
    #ignoring NAs.
calculate_diff <- function(x){
  x <- as.numeric(x)
  x[which(!is.na(x))] <- c(NA, diff(na.omit(x)))
  x
}
#source: https://stackoverflow.com/questions/68496968/find-the-difference-of-a-column-ignoring-nas-find-difference-only-of-numbers

#apply change function
ht_changes <- cbind(growth_list_init$static$id, as.data.frame(t(apply(growth_list_init$ht_in, 1, calculate_diff))))
colnames(ht_changes) <- c("id", paste0("ht_chg_", 1:11))

#get all observations with negative height changes; match in heights for context
neg_ht_changes <- 
  left_join(
    ht_changes %>% filter_at(vars(-id), any_vars(. < 0)),
    cbind(data.frame(id = growth_list_init$static$id), growth_list_init$ht_in), by = "id")

#select for observations with negative changes with a magnitude greater than 1 inch.
neg_ht_changes_1 <- neg_ht_changes %>% filter_at(vars(-id), any_vars(. < -1))
#get ids of observations with negative height changes with a magnitude greater than 1 inch.
problem_ids <- neg_ht_changes_1$id
problem_id_idxs <- which(growth_list_init$static$id %in% problem_ids)

growth_list <- lapply(growth_list_init, function(x){x[-problem_id_idxs,]})
#exclude all observations with negative height changes greater than 1 inch.
```

##Do unit conversions
```{r}
#get all height measurements in cm
#apply function across dataframe columns, convert resulting matrix to dataframe,
  #then assign to list item
growth_list$ht_cm <- as.data.frame(apply(growth_list$ht_in, 2, conv_unit, "in", "cm"))

growth_list$wt_kg <- as.data.frame(apply(growth_list$wt_lbs, 2, conv_unit, "lbs", "kg"))

names(growth_list)

colnames(growth_list$ht_in) <- paste0("ht_in_", 1:11)
colnames(growth_list$ht_cm) <- paste0("ht_cm_", 1:11)
colnames(growth_list$wt_kg) <- paste0("wt_kg_", 1:11)
colnames(growth_list$wt_lbs) <- paste0("wt_lbs_", 1:11)
```

________________________________________________________________________________


# Standardize health values

##Apply CDC package (https://github.com/CDC-DNPAO/CDCAnthro)
```{r}
#specific function for working with these data: use cdc function "cdcanthro"
#for a point in time, producing "z-scores, percentiles, and other metrics for 
#weight, height, and BMI based on the 2000 CDC growth charts"
#input: my_num, integer in [1,11]
#output: data table (see cdcanthro documentation)
use_cdcanthro <- function(my_num){
  cdcanthro(cbind(data.frame(id = growth_list$static$id), 
                  data.frame(sex = growth_list$static$sex), 
                  cdcage = growth_list$cdcage[,my_num], 
                  wt_kg = growth_list$wt_kg[,my_num], 
                  ht_cm = growth_list$ht_cm[,my_num],
                  BMI = growth_list$BMI[,my_num]),
            cdcage,
            wt_kg,
            ht_cm,
            BMI,
            all = TRUE
          )
}

#get data tables of cdcanthro output for each measurement point
cdc_result_list <- lapply(1:11, use_cdcanthro)
```

cdcanthro output:
- waz, haz, bmiz: CDC –for-age z-scores for Weight, Height, and BMI. BMIz is based on 2000 CDC growth charts (non-obese children) and extended BMIz (obese children)
- bmip95: BMI expressed as percentage of 95th percentile; 120 percent is lower threshold for severe obesity

Note: if other variables are needed, they can easily be pulled from cdc_result_list using function 'consolidate_dfs'

##Function to pull relevant columns from cdc algorithm output into our data
```{r}
#function: consolidate identically-named columns in each dataframe in a list
#into a single dataframe, labeled with sequentially increasing integers [1,max]
#inputs: 
  #df_list: list of dataframes (or other arrays) 
  #my_colname: string that's a column name that appears in each dataframe in df_list
#output:
  #var_df: dataframe which compiles each instance of my_colname in df_list
consolidate_dfs <- function(df_list, my_colname){
  #retrieve only relevant columns
  var_list <- lapply(df_list, function(df){df[[my_colname]]})
  #check for dataframes without the relevant column; issue warning.
  nulls_present <- as.logical(lapply(var_list,is.null))
  if(any(nulls_present)){
    stop(paste0("Column '", my_colname, "' doesn't appear in list element(s) ", 
                paste(seq(1,length(df_list))[nulls_present],collapse = ", ")))
  }
  #compile columns into dataframe
  var_df <- data.frame(var_list)
  #label columns in new dataframe
  colnames(var_df) <- paste0(my_colname, "_", 1:ncol(var_df))
  var_df
}
```

##Pull relevant columns from cdc algorithm output into our data
```{r}
growth_list$cdc_bmiz <- consolidate_dfs(cdc_result_list, "bmiz")
growth_list$cdc_bmip <- consolidate_dfs(cdc_result_list, "bmip")
growth_list$cdc_waz <- consolidate_dfs(cdc_result_list, "waz")
growth_list$cdc_wap <- consolidate_dfs(cdc_result_list, "wap")
growth_list$cdc_haz <- consolidate_dfs(cdc_result_list, "haz")
growth_list$cdc_hap <- consolidate_dfs(cdc_result_list, "hap")
growth_list$cdc_perc_median <- consolidate_dfs(cdc_result_list, "perc_median")
growth_list$cdc_bmip95 <- consolidate_dfs(cdc_result_list, "bmip95")

#remove large list if needed to save memory
#rm(cdc_result_list)
```

________________________________________________________________________________


# Check for Data Quality Issues: Extreme Values

##Check for duplicates
```{r}
any(duplicated(growth_list$static$id))
#none
```

##Function to check for unusual trends
```{r}
#outlier flags present in SAS code but not R package. reconstructed here.
#source: https://www.cdc.gov/nccdphp/dnpao/growthcharts/resources/sas.htm

#function to check whether values are within a given plausibility range
  #inputs:
    #z_score: the value to check: in this context, a z_score for weight, height, 
      #or bmi. a numeric value.
    #lower: the lower plausible bound. a numeric value.
    #upper: the upper plausible bound. a numeric value.
  #output: biv, a value (-1 for low, 0 for plausible, +1 for high)
biv_flag <- function(z_score, lower, upper){
  if(lower >= upper){
    stop("lower bound for plausible values must be less than upper bound")
  }
  if(is.na(z_score)){return(NA)}
  if(!is.numeric(c(z_score,lower,upper))){
    stop("all input values must be numeric")
  }
  biv <- if(z_score < lower){-1} else if (z_score > upper){1} else {0}
  return(biv)
}
```

##Apply function: check for unusual values with given CDC cutoffs
```{r}
#Weight-for-age for children aged from 0 to < 240 months: bivwt
  #Cutoff for extreme z-scores: 
    #Low (flag -1): < -5
    #High (flag +1): > 8
bivwt <- as.data.frame(apply(growth_list$cdc_waz, c(1,2), biv_flag, -5, 8))
colnames(bivwt) <- paste0("bivwt_", 1:11)
# growth_list$cdc_bivwt <- bivwt

#Height-for-age for children aged from 0 to < 240 months: bivht
  #Cutoff for extreme z-scores: 
    #Low (flag -1): < -5
    #High (flag +1): > 4
bivht <- as.data.frame(apply(growth_list$cdc_haz, c(1,2), biv_flag, -5, 4))
colnames(bivht) <- paste0("bivht_", 1:11)
# growth_list$cdc_bivht <- bivht

#BMI-for-age for children aged 24 to < 240 months: bivbmi
  #Cutoff for extreme z-scores: 
    #Low (flag -1): < -4
    #High (flag +1): > 8
bivbmi <- as.data.frame(apply(growth_list$cdc_bmiz, c(1,2), biv_flag, -4, 8))
colnames(bivbmi) <- paste0("bivbmi_", 1:11)
# growth_list$cdc_bivbmi <- bivbmi
```

##Separate extreme values
```{r}
#get observations with extreme values for weight
ext_wt <- 
  cbind(id = growth_list$static$id, bivwt) %>% filter_at(vars(-id), any_vars(. != 0))
dim(ext_wt)

#get observations with extreme values for height
ext_ht <- 
  left_join(
    cbind(id = growth_list$static$id, bivht) %>% filter_at(vars(-id), any_vars(. != 0)),
    cbind(id = growth_list$static$id, growth_list$ht_in), by = "id")
dim(ext_ht)

#get observations with extreme values for bmi
ext_bmi <- 
  cbind(id = growth_list$static$id, bivbmi) %>% filter_at(vars(-id), any_vars(. != 0))
dim(ext_bmi)

#get ids and id indices for observations with extreme values
ext_ids <- c(ext_wt$id, ext_ht$id, ext_bmi$id)
ext_id_idxs <- which(growth_list$static$id %in% ext_ids)
length(ext_id_idxs) #number of observations removed

#Remove observations with CDC-flagged extreme values for weight, height, and BMI.
growth_list1.1 <- lapply(growth_list, function(x){x[-ext_id_idxs,]})
```

##check for rare values in static features
```{r}
summary(growth_list1.1$static)
#identify: few values for indigenous American in `ethnic`, plus 97 NAs.

few_ethnic_idxs <- which(growth_list1.1$static$ethnic == 5 | is.na(growth_list1.1$static$ethnic))
length(few_ethnic_idxs) #number of observations removed

growth_list2 <- lapply(growth_list1.1, function(x){x[-few_ethnic_idxs,]})
```
Remove observations with missing values for ethnicity or ethnicity = indigenous American (too few observations).

________________________________________________________________________________


# Check for Data Quality Issues: Missing Data

##Address missing data
```{r}
##function to get proportion of missing values along different axes
#inputs:
  #df: dataframe
  #direction: integer 1 (horizontal), 2 (vertical), 0 (overall)
#outputs:
  #if horizontal or vertical: dataframe with proportions of missing values, labeled.
  #if overall: single float with overall proportion of missing values
get_prop_missing <- function(df, direction){
  if(!(direction %in% c(1,2,0))){stop("Invalid direction. Try: 1 (horizontal), 2 (vertical), or 0 (overall).")}
  df_na_logic <- apply(df, 2, is.na)
  if(direction == 0){return(mean(df_na_logic))}
  prop_na <- apply(df_na_logic,direction,mean)
  if(direction == 2){as.data.frame(prop_na)}else(cbind(data.frame(id = growth_list2$static$id),prop_na))
}

get_prop_missing(growth_list2$ht_in, 2)
#get_prop_missing(growth_list2$ht_in, 1) 
  #individual-level data; can uncomment and run in context
get_prop_missing(growth_list2$ht_in, 0)
```

##Visualize missing data
```{r}
SCHOOL_TERMS = c('K', 'K-S', '1', '1-S', '2', '2-S', '3', '3-S', '4', '4-S', '5')

##function to get bar charts with proportion of missing data over time
#input: string variable name (no number attached)
#output: ggplot bar graph
missing_bar <- function(varname, varstring){
  #error message for improper input
  if(!(varname %in% names(growth_list2))){
    stop(paste0(varname, " not a column name in growth_list2"))}
  #run missing data function
  df <- get_prop_missing(growth_list2[[varname]], 2)
  df_named <- df %>% mutate(var = rownames(df),
                            school_terms = SCHOOL_TERMS) #retrieve row names as col for ggplot
  
  ggplot(df_named) + aes(x = fct_inorder(school_terms), y = prop_na, 
                 label = scales::percent(round(prop_na,3))) +
  geom_col(fill = "#21918c") +
  theme_bw() + 
  geom_text(position = position_dodge(width = .9),    # move to center of bars
              vjust = -0.5,    # nudge above top of bar
              size = 3) +
  scale_y_continuous(labels = scales::percent, limits = c(0,1)) +
  labs(y = "Proportion of Observations Missing", 
       x = "Observation Point",
       title = paste("Missing data for", varstring, "variable"))
}
#https://stackoverflow.com/questions/40249943/adding-percentage-labels-to-a-bar-chart-in-ggplot2
missing_bar("ht_in", "height (in.)")

#how many kids are at each "percent missing" proportion?
# get_prop_missing(growth_list2$ht_in, 1) %>%
#   group_by(prop_na) %>% summarize(count_NAs = n())

#how many are missing at least X proportion?

##function to get proportion of population missing observations for a given variable
#input:
  #varname: string that is an element in growth_list2
  #prop: float in [0,1]
  #string: boolean for string vs. float return
#output: descriptive string: "X% of kids are missing at least 'prop'*100% of observations
  #for variable 'varname', or float X if string = FALSE
missing_prop<- function(varname, varstring, prop, string = TRUE){
  #error messages for improper input
  if(!(varname %in% names(growth_list2))){
    stop(paste0(varname, " not a column name in growth_list2"))}
  if(prop < 0 | prop > 1){stop("prop must be in [0,1]")}
  num <- get_prop_missing(growth_list2[[varname]], 1) %>%
           summarize(missing_x = mean(prop_na >= prop))
  #output string
  ifelse(string, 
         paste0(round(num[1,1], 3) * 100, "% of kids are missing at least ", 
         round(prop, 3) * 100, "% of observations for the ", varstring, " variable"),
         num[1,1])
}
missing_prop("ht_in", "height (in.)", 0.5)
missing_prop("ht_in", "height (in.)", 1/11, FALSE)

#how many observations missing overall?
get_prop_missing(growth_list2$ht_in, 0)
```

#Function to integrate previous missing-data-visualization functions
```{r}
#percentage of each variable that's missing: height, weight, BMI, school

##function integrating missing_bar and missing_prop
#input: varname: string that is the name of an element in growth_list2
#output: ggplot with caption
missing_bar_w_capt <- function(varname, varstring = varname){
  missing_bar(varname,varstring) +
    labs(caption = paste0(missing_prop(varname, varstring, 0.5), "; ", 
                         round(missing_prop(varname, varstring, 1/11, FALSE), 3) * 100, 
                        "% are missing at least one."))
}
```

#Visualize missing data
```{r, fig.dim = c(7.2,5.2)}
missing_bar_w_capt("ht_in", "height (in.)")
missing_bar_w_capt("wt_lbs", "weight (lbs)")
missing_bar_w_capt("cdc_bmiz")
missing_bar_w_capt("school")
```

________________________________________________________________________________


# Join Supplementary Datasets to Main Dataset

#Join census tract codes to schools
```{r}
#join each column of school on school code to tract
#create new dataframe list object with tracts
tracts_simple <- tracts %>% select(Number, `TRACT CODE`) %>% 
  rename(school_num = Number, tract_code = `TRACT CODE`) %>%
  mutate(school_num = factor(school_num))

#using an ugly loop here to match schools to census tract codes
#have to run 11 times for 11 left joins, with appropriate renaming and indexing
tract <- data.frame(tract = numeric(nrow(growth_list2$static)))
for(i in 1:11){
  colname <- paste0("school_", i)
  tract[,i] <-
  left_join(growth_list2$school %>% select(as.name(colname)) %>%
              rename(school_num = as.name(colname)), 
            tracts_simple, by = "school_num") %>%
    select(tract_code)
  rm(colname)
}
colnames(tract) <- paste0("tract_", 1:11)

growth_list2$tract <- tract
```

#Join COI data
```{r}
COI_simple <- COI %>% select(census_tract, r_COI_nat) %>%
  rename(tract = census_tract,
         COI_nat = r_COI_nat) %>%
  mutate(tract = as.character(tract))
#summary(COI_simple)

#ugly loop here to match kids to COI on census tract
COI_df <- data.frame(COI_nat = character(nrow(growth_list2$static)))

for(i in 1:11){
  colname <- paste0("tract_", i)
  
  COI_df[,i] <- 
    left_join(growth_list2$tract %>% select(as.name(colname)) %>%
                rename(tract = as.name(colname)), 
              COI_simple, by = "tract") %>%
    select(COI_nat)
  rm(colname)
}
colnames(COI_df) <- paste0("COI_nat_", 1:11)

growth_list2$COI_nat <- COI_df
```

#Join ALAN (DMSP) data
```{r}
#get versions of dataframes for matching
DMSP_simple <- DMSP %>% select(TRACTCE, YEAR, MEAN) %>%
  mutate(TRACTCE = as.character(TRACTCE),
         YEAR = as.character(YEAR)) %>%
  rename(tract = TRACTCE,
         year = YEAR,
         DMSP_mean = MEAN)

growth_years <- apply(growth_list2$dcoll, 2, function(x){as.character(year(x))})
colnames(growth_years) <- paste0("year_", 1:11)
growth_match <- cbind(growth_years, growth_list2$tract)

#ugly loop here to match kids to light level on census tract and year
DMSP_df <- data.frame(DMSP_mean = numeric(nrow(growth_list2$static)))

for(i in 1:11){
  tract_colname <- paste0("tract_", i)
  year_colname <- paste0("year_", i)
  
  DMSP_df[,i] <- 
    left_join(growth_match %>% select(as.name(tract_colname), as.name(year_colname)) %>%
                rename(tract = as.name(tract_colname),
                       year = as.name(year_colname)), 
              DMSP_simple, by = c("tract", "year")) %>%
    select(DMSP_mean)
  rm(tract_colname, year_colname)
}
colnames(DMSP_df) <- paste0("DMSP_mean_", 1:11)

growth_list2$DMSP_mean <- DMSP_df

#check missingness
tract_col_NAs
(DMSP_mean_col_NAs <-apply(growth_list2$DMSP_mean, 2, is_na_sum))
```

________________________________________________________________________________


# Output Cleaned Data

```{r}
#"re-flatten" data, ready for output
growth_flat <- bind_cols(growth_list2)
```

```{r}
#make codebook
codebook <- 
data.frame(variable = c(colnames(growth_list2$static), names(growth_list2[-1])),
           label = c("Student ID",
                     "Ethnicity (1 = white, 2 = Black, 3 = Hispanic, 4 = Asian)",
                     "Gender (1 = male, 2 = female)",
                     "Date of birth",
                     "Date collected (times 1-11)",
                     "Height (in.) (times 1-11)",
                     "Weight (lbs.) (times 1-11)",
                     "BMI calculated (times 1-11)",
                     "School code (times 1-11)",
                     "Age in months (times 1-11)",
                     "Height (cm.) (times 1-11)",
                     "Weight (kg.) (times 1-11)",
                     "BMI z-score, CDC for-age (times 1-11)",
                     "BMI percentile, CDC for-age (times 1-11)",
                     "Weight z-score, CDC for-age (times 1-11)",
                     "Weight percentile, CDC for-age (times 1-11)",
                     "Height z-score, CDC for-age (times 1-11)",
                     "Height percentile, CDC for-age (times 1-11)",
                     "BMI as percentage of median, CDC for-age (times 1-11)",
                     "BMI as percentage of 95th percentile, CDC for-age (times 1-11)",
                     "Census Tract GEOID code (times 1-11)",
                     "COI score for census tract (normalized by national ranking) (times 1-11)",
                     "Mean of annual measured light for census tract (times 1-11)"))
```

```{r}
#check number of complete records
sum(apply(growth_flat, 1, function(x){!any(is.na(x))}))
```

```{r}
#write data and codebook to csv
write.csv(growth_flat, file = "../data_directory/growth.csv")
write.csv(codebook, "growth_codebook.csv")
```


